Management induced changes of soil organic carbon on global croplands

### 2020-12-14 SOCbudgetPaper/methods.Rmd
update methods

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol. 4 ([@calvo_buendia_ipcc_2019]). We additionally describe how we configured and extended the steady-state method (for model code see [@karstens_mrsoil_2020]). In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines ([@eggleston_ipcc_2006], [@calvo_buendia_ipcc_2019]). 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see [@bodirsky_mrcommons_2020]). In Sect. \@ref(sec:scenarios) we define the scenarios used to complement our historic model results.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model ([@parton_analysis_1987]). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

@@

  ```{=latex}
    \begin{table*}[h]
    \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
    \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.40\textwidth}}
    \tophline
    \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
    \middlehline
    \multirow{3}{*}{Cropland} & above-ground residues & \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                        \cite{faostat_faostat_2016}, \\
                                                                        \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                        \cite{weindl_livestock_2017} \end{minipage}} & 
                                                        \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                        LG:C generic values according to Table 5.5B\\
                                                                        in \cite{calvo_buendia_ipcc_2019},\\
                                                                        crop-type specific N:C-ratios from
                                                                        \cite{bodirsky_mrcommons_2020}\end{minipage}} \\
                              & below-ground residues &  &  \\
                              & manure                &  &  \\ 
                              \hline
    Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                              & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                  \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
   \bottomhline
   \end{tabular}
   \label{tab:datasourceinputs}
   \belowtable{}
   \end{table*}
```

@@

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm 
                                     & \cite{hengl_soilgrids250m_2017} \\
                      \cline{2-4}

@@

Note that the absolute changes $\Delta SOC$ can be also interpreted as the SOC debt ([@sanderman_soil_2017]) due to human cropping activities; whereas relative changes $F^{\mathrm{SCF}}$ can be considered stock change factors as defined within the IPCC guidelines of 2006 [@eggleston_ipcc_2006]. Moreover, $\Delta SOC$ is equivalent to the negated cumulative SOC component of human land-use change emissions ([@pugh_simulated_2015]).

@@

Additionally to the steady-state method [@calvo_buendia_ipcc_2019] and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines ([@eggleston_ipcc_2006], [@calvo_buendia_ipcc_2019]). Here, stocks are calculated via stock change factors ($F^{\mathrm{SCF}}$) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of $F^{\mathrm{SCF}}$ are differentiated by different crops, management and input systems (here summarized under $m$) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, estimates of $F^{\mathrm{SCF}}$ vary for different climatic zones ($c$) specified by the IPCC (see Fig. \@ref(fig:CLIMzone)). The actual $SOC$ stocks are thus calculated based on a given reference stock $SOC^{\mathrm{ref}}$ by

@@

with $T_{c,i}$ being the translation matrix for grid cells $i$ into corresponding climate zones $c$. For this analysis, we use the default $F^{\mathrm{SCF}}$ from the Tier 1 method of [@eggleston_ipcc_2006] and [@calvo_buendia_ipcc_2019] as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016] (see FAO2LUH2MAG_croptypes.csv in [@bodirsky_mrcommons_2020]). By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL [@schaphoff_lpjml4_2018] as well as half-degree cropland patterns (see Sect. \@ref(sec:landuse)). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production [@faostat_faostat_2016]. By using physical cropland areas in combination with harvested areas, we account for multiple cropping systems as well as for fallow land.

@@

Following the IPCC guidelines, we split the harvest index into a production ($HI^{\mathrm{prod}}$) and an area dependent ($HI^{\mathrm{area}}$) fraction [@eggleston_ipcc_2006]. Deviating from [@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following [@smil_nitrogen_1999]) we assume a burn share of $25\%$ for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_ipcc_2006].

@@

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon resp. nitrogen to dry matter (c/dm, n/dm) ratios (see \@ref(tab:c2dm)).

@@

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization and parameter choices, we perform a set of scenario runs. In the following sections we outline name and idea of these scenarios (for technical implementation see [@karstens_mrsoil_2020]).

@@

As outlined in Sect. \@ref(sec:initsoc) we assume landuse-type specific steady-state SOC stocks for the start year 1901 followed by a long spin-up period of 74 years ((\textit{Initial-lu})). As some SOC compartments decay over very long timescales, the initialisation setting might strongly effect the overall outcome of SOC stocks and changes. Thus we conduct counterfactual scenario \textit{Initial-natveg}, initializing the start year with the steady-state SOC under potential natural vegetation for all land-use types. We additionally combined the \textit{Initial-natveg} scenario with the (\textit{constManagement} scenario.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines ([@calvo_buendia_ipcc_2019]), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas LG:C ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, N:C ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table \@ref(tab:scenlitterpnv).

>>>>

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol. 4 ([@calvo_buendia_ipcc_2019]). We additionally describe how we configured and extended the steady-state method (for model code see [@karstens_mrsoil_2020]). In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]). 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see [@bodirsky_mrcommons_2020]). In Sect. \@ref(sec:scenarios) we define the scenarios used to complement our historic model results.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model [@parton_analysis_1987]). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

@@

  ```{=latex}
    \begin{table*}[h]
    \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
    \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.25\textwidth}}
    \tophline
    \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
    \middlehline
    \multirow{3}{*}{Cropland} & above-ground residues & \multirow{3}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                        \cite{faostat_faostat_2016}, \\
                                                                        \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                        \cite{weindl_livestock_2017} \end{minipage}} & 
                                                        \multirow{2}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                        LG:C generic values according to Table 5.5B\\
                                                                        in \cite{calvo_buendia_ipcc_2019},\\
                                                                        crop-type specific N:C-ratios from
                                                                        \cite{bodirsky_mrcommons_2020}\end{minipage}} \\ \cline{2-2}
                              & below-ground residues &  &  \\ \cline{2-2}
                              & manure                &  & Table 5.5C in \cite{calvo_buendia_ipcc_2019} \\ 
                              \hline
    Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                              & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                  \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
   \bottomhline
   \end{tabular}
   \label{tab:datasourceinputs}
   \belowtable{}
   \end{table*}
```

@@

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm 
                                     & \cite{hengl_soilgrids250m:_2017} \\
                      \cline{2-4}

@@

Note that the absolute changes $\Delta SOC$ can be also interpreted as the SOC debt ([@sanderman_soil_2017]) due to human cropping activities; whereas relative changes $F^{\mathrm{SCF}}$ can be considered stock change factors as defined within the IPCC guidelines of 2006 [@eggleston_hs_ipcc_2006]. The SOC debt is equivalent to the negated cumulative SOC component of  of human land use emission ([@pugh_simulated_2015]).

@@

Additionally to the steady-state method [@calvo_buendia_ipcc_2019] and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]). Here, stocks are calculated via stock change factors ($F^{\mathrm{SCF}}$) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of $F^{\mathrm{SCF}}$ are differentiated by different crops, management and input systems (here summarized under $m$) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, estimates of $F^{\mathrm{SCF}}$ vary for different climatic zones ($c$) specified by the IPCC (see Fig. \@ref(append:climatemap)). The actual $SOC$ stocks are thus calculated based on a given reference stock $SOC^{\mathrm{ref}}$ by

@@

with $T_{c,i}$ being the translation matrix for grid cells $i$ into corresponding climate zones $c$. For this analysis, we use the default $F^{\mathrm{SCF}}$ from the Tier 1 method of [@eggleston_hs_ipcc_2006] and [@calvo_buendia_ipcc_2019] as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016] (see FAO2LUH2MAG_croptypes.csv in [@mrcommons]). By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL [@schaphoff_lpjml4_2018] as well as half-degree cropland patterns (see Sect. \@ref(sec:landuse)). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historic FAOSTAT production [@faostat_faostat_2016]. By using physical cropland areas in combination with harvested areas, we account for multiple cropping systems as well as for fallow land.

@@

Following the IPCC guidelines, we split the harvest index into a production ($HI^{\mathrm{prod}}$) and an area dependent ($HI^{\mathrm{area}}$) fraction [@eggleston_hs_ipcc_2006]. Deviating from [@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following [@smil_nitrogen_1999]) we assume a burn share of $25\%$ for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_hs_ipcc_2006].

@@

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon resp. nitrogen to dry matter (c/dm, n/dm) ratios (see \@ref(append:Tablec2dm)).

@@

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization and parameter choices, we perform a set of scenario runs. In the following sections we outline name and idea of these scenarios (for technical implementation see [@mrsoil_karstens_2020]).

@@

As outlined in Sect. @ref(sec:initsoc) we assume landuse-type specific steady-state SOC stocks for the start year 1901 followed by a long spin-up period of 74 years ((\textit{Initial-lu})). As some SOC compartments decay over very long timescales, the initialisation setting might strongly effect the overall outcome of SOC stocks and changes. Thus we conduct counterfactual scenario \textit{Initial-natveg}, initializing the start year with the steady-state SOC under potential natural vegetation for all land-use types. We additionally combined the \textit{Initial-natveg} scenario with the (\textit{constManagement} scenario.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines ([@calvo_buendia_ipcc_2019]), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas LG:C ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, N:C ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table @ref(tab:scenlitterpnv).
